---
title: "Binary outcomes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Binary outcomes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Example

The example uses the `breastcancer` data set from the risks package and compares the risk of death (a binary variable in this case) by categories of cancer stage.

```{r example, message = FALSE}
library(rifttable)
library(dplyr)
data(breastcancer, package = "risks")

tibble::tribble(
  ~label,                      ~type,
  "**Absolute estimates**",    "",
  "Observations",              "total",
  "Outcomes",                  "outcomes",
  "Outcomes/Total",            "outcomes/total",
  "Cases/Controls",            "cases/controls",
  "Risk",                      "risk",
  "Risk (95% CI)",             "risk (ci)",
  "Outcomes (Risk)",           "outcomes (risk)",
  "Outcomes/Total (Risk)",     "outcomes/total (risk)",
  "",                          "",
  "**Comparative estimates**", "",
  "Risk ratio (95% CI)",       "rr",
  "Risk difference (95% CI)",  "rd",
  "Odds ratio (95% CI)",       "or") %>%
  mutate(
    exposure = "stage",
    outcome = "death") %>%
  rifttable(
    data = breastcancer,
    overall = TRUE) %>%
  rt_gt()  # Formatted output
```


# Absolute estimates per exposure category

`type` | Description | Options (`arguments = `)
-------+-------------+--------------
`"cases/controls"` | Cases and non-cases (events and non-events); useful for case-control studies.
`"outcomes"` | Outcome count.
`"outcomes (risk)"` | A combination: Outcomes followed by risk in parentheses.
`"outcomes/total (risk)"` | A combination: Outcomes slash total followed by risk in parentheses.
`"risk"` | Risk (or prevalence), calculated as a proportion, *i.e.*, outcomes divided by number of observations. Change between display as proportion or percent using the parameter `risk_percent` of `rifttable()`.
`"risk (ci)"` | Risk with confidence interval (default: 95%): Wilson score interval for binomial proportions, see `rifttable::scoreci()`.


# Comparative estimates with confidence intervals

`type` | Description | Options (`arguments = `)
-------+-------------+--------------
`"irr"` | Incidence rate ratio for count outcomes from Poisson regression model, with confidence interval (default: 95%).
`"or"` | Odds ratio from logistic regression, with confidence interval (default: 95%).
`"rd"` | Risk difference (or prevalence difference) from risks::riskdiff(), with confidence interval (default: 95%). | `list(approach = "margstd_boot", bootrepeats = 2000)` to request model fitting via marginal standardization with 2000 bootstraps.
`"rr"` | Risk ratio (or prevalence ratio) from `risks::riskratio()`, with confidence interval (default: 95%). | See `"rd"`.
