---
title: "Frequently Asked Questions"
output: 
  rmarkdown::html_vignette:
    df_print: kable
vignette: >
  %\VignetteIndexEntry{Frequently Asked Questions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Basic example

Most subsequent examples build on this simple rifttable:

```{r basic, message = FALSE}
library(rifttable)
library(dplyr)  # for data management, e.g., mutate()
library(tibble)  # for constructing a tibble, e.g. via tribble()
data(breastcancer, package = "risks")

design <- tribble(
  ~label,                       ~type,                   ~stratum,         
  "**Overall**",                "",                      "",               
  "  Deaths/N",                 "outcomes/total",        c("Low", "High"), 
  "  Risk",                     "risk",                  c("Low", "High"), 
  "  Risk ratio (95% CI)",      "rr",                    c("Low", "High"), 
  "  Risk difference (95% CI)", "rd",                    c("Low", "High"), 
  "",                           "",                      "",               
  "**Low hormone receptor**",   "",                      "",               
  "  Deaths/N (Risk)",          "outcomes/total (risk)", "Low",           
  "  Risk difference (95% CI)", "rd",                    "Low",           
  "**High hormone receptor**",  "",                      "",               
  "  Deaths/N (Risk)",          "outcomes/total (risk)", "High",
  "  Risk difference (95% CI)", "rd",                    "High") %>%
  mutate(
    exposure = "stage",
    outcome = "death",
    effect_modifier = "receptor")

rifttable(
  design = design,
  data = breastcancer) %>%
  rt_gt()  # obtain formatted output
```


# Why do I get an error?

R's error messages can be frustrating. When using rifttable, these are the typical sources of errors:

* Clerical errors in variable names and arguments. There is no magic here except double-checking the code.
* Missing data. See below: [How do I handle missing data?](#how-do-i-handle-missing-data).
* Discrepancy between estimator (`type`) and the data. For example, `type = "mean"` will not work on a categorical (factor) variable.
* Models that fail to converge. For example, one may be trying to estimate a risk ratio with 0 outcomes in the reference category, or adjusting for 20 covariates in a Cox model with 5 events overall. (Sometimes such attempts will "just" return warning messages -- it is still worth rethinking the modeling strategy.)

To identify where an error is coming from, start simple. Comment out all but one line of the `design`, putting `#` at the beginning of the line. Start with a line that gives basic descriptive data, such as `type = "total"`, `type = "outcomes"` or `type = "events/time"`, and re-run `rifttable()`. Then add more lines with descriptive estimators, one by one. At the end, add lines that fit models, such as `type = "hr"`.


# What is the `design`?

The `design` that the `rifttable()` function takes as input is simply a dataset that defines how the table should look like when `rifttable()` has processed the `data`.

The `design` can be constructed in many different ways. All lead to the same table:

1. **A dataset (tibble) defined using `tribble()`**

    ```{r}
    design1 <- tribble(
      ~label,   ~exposure, ~outcome, ~type,
      "N",      "stage",   "death",  "total",
      "Deaths", "stage",   "death",  "outcomes")
    design1
    rifttable(
      design = design1,
      data = breastcancer) %>%
      rt_gt()
    ```

2. **A dataset (tibble) defined using `tibble()`**

    ```{r}
    design2 <- tibble(
        label = c("N", "Deaths"),
        exposure = "stage",
        outcome = "death", 
        type = c("total", "outcomes"))
    design2
    rifttable(
      design = design2,
      data = breastcancer) %>%
      rt_gt()
    ```

3. **Concatenating tibbles, then editing with `mutate()`**

    ```{r}
    design3 <- bind_rows(
      tibble(  # row 1
        label = "N",
        type = "total"),
      tibble(  # row 2
        label = "Deaths",
        type = "outcomes")) %>%
      mutate(  # elements that are the same for all rows
        exposure = "stage",
        outcome = "death")
    design3
    rifttable(
      design = design3,
      data = breastcancer) %>%
      rt_gt()
    ```


4. **For descriptive tables: Use `table1_design()`**

    ```{r}
    design4 <- breastcancer %>%
      table1_design(
        death,  # the total count will automatically be included
        by = stage)
    design4
    rifttable(
      design = design4,
      data = breastcancer) %>%
      rt_gt()
    ```
    
    See a separate overview about [a descriptive Table 1](table1.html).

5. **External datasets**

    The `design` could even be written out in an external dataset that can be
    loaded with `readr::read_csv()` (for CSV files) or `readxl::read_excel()`
    (for Excel sheets).


# How do I handle missing data?

rifttable tries to make as few assumptions as possible about how the user wants to treat missing data.

* Missing values in **exposure**: By default, missing vales (`NA`) in the exposure are displayed as a separate exposure category for descriptive statistics (e.g., `type = "total"` or `type = "mean"`). They are omitted by comparative estimators (e.g., regression models). To change this behavior, call `rifttable()` with the argument `exposure_levels = "nona"`.
* Missing values in **outcome**: By default, descriptive statistics will be missing (e.g., results will be `--` or `NA`), and regression models will use the non-missing observations. To exclude observations with missing outcome values altogether, add `na_rm = TRUE` to (specific rows of) the `design`.
* Missing values in **confounders**: Applies only to regression models, which typically exclude observations with missing values.
* Missing values in the **effect modifier**: Stratified and joint models are only shown for the specified `stratum` of the `effect_modifier`. To include observations with the missing effect modifier, add `NA` to the requested stratum in the `design`, e.g., `effect_modifier = "bmi", stratum = c("<25", NA)`.


# How do I add overall statistics?

Use the `overall` argument to show descriptive data for the entire `data` set.
Inferential estimators showing comparisons between exposure categories will be blank there.

```{r overall}
rifttable(
  design = design,
  data = breastcancer, 
  overall = TRUE) %>%
  rt_gt()  # obtain formatted output
```


# How do I test for trend?

Instead of *testing* a null hypothesis about a trend, rifttable proposes 
*estimating* the difference in the outcome for a one-unit higher exposure. 
This is also called a linear slope. Here, we estimate the risk associated for 
stage that is one category higher.

```{r trend}
rifttable(
  design = design %>%
    mutate(trend = "stage_numeric"),
  data = breastcancer %>%
    mutate(stage_numeric = as.numeric(stage))) %>%
  rt_gt()  # obtain formatted output
```


# How do I show multiple exposures in the same table?

Our simple toy dataset just has one exposure variable. For demonstration,
we just create a second variable, with two categories, "Level 1" and "Level 2,"
which is a simplified combination of the `stage` and `receptor` variables.

We will flip the table `layout` from `"rows"` (the default) to `"cols"` and 
concatenate two rifttables. We also need to give our new `exposure2` variable 
the same label as `stage` to make sure results appear in 
the same column.


```{r multiple}
breastcancer_2exposures <- breastcancer %>%
  mutate(
    exposure2 = case_when(
      stage == "Stage I" | 
        (stage == "Stage II" & receptor == "High") ~ 
        "Level 1",
      stage == "Stage III" | 
        (stage == "Stage II" & receptor == "Low") ~ 
        "Level 2"))

attr(breastcancer_2exposures$exposure2, which = "label") <- "Exposure"
attr(breastcancer_2exposures$stage, which = "label") <- "Exposure"

bind_rows(
  design %>%
    mutate(exposure = "exposure2") %>%
    slice(2:5) %>%
    rifttable(
      data = breastcancer_2exposures, 
      layout = "cols"),
  design %>%
    slice(2:5) %>%
    rifttable(
      data = breastcancer_2exposures, 
      layout = "cols")) %>%
  rt_gt()  # obtain formatted output
```


# How do I change how results are rounded?

By default, difference measures are being rounded to 2 decimal digits (`0.01`), such as `type = "diff"`, the mean difference, or `type = "quantreg"`, the median difference. The same goes for risk measures, such as `type = "risk"`, unless shown as percentage points. Ratio measures are also shown with 2 decimal digits, such as `type = "hr"`, the hazard ratio, or `type = "fold"`, a ratio of arithmetric means. 

Rounding can be changed by setting the `rifttable()` arguments `diff_digits`, `risk_digits`, and `ratio_digits` globally for the entire table. 


```{r rounding}
design <- tribble(
  ~label,                     ~type,
  "Deaths/N",                 "outcomes/total", 
  "Risk",                     "risk",          
  "Risk ratio (95% CI)",      "rr", 
  "Odds ratio (95% CI)",      "or", 
  "Risk difference (95% CI)", "rd") %>%
  mutate(
    exposure = "stage",
    outcome = "death",
    effect_modifier = "receptor")

rifttable(
  design = design,
  data = breastcancer,
  ratio_digits = 3,  # Many digits for ratios
  risk_digits = 1) %>%  # Fewer digits for risks
  rt_gt()  # obtain formatted output
```

As can be seen, ratios > 3 are still shown with 1 fewer decimal, and ratios > 10 are shown with 2 fewer decimals ([Wilcox, *Epidemiology* 2004](https://doi.org/10.1097/01.ede.0000101026.08873.14) motivates why). To disable such additional rounding of extremely high ratios:

```{r rounding_decrease}
rifttable(
  design = design,
  data = breastcancer,
  ratio_digits = 3,
  ratio_digits_decrease = NULL,  # Do not round high ratios more
  risk_digits = 1) %>%
  rt_gt()  # obtain formatted output
```

Additionally, rounding can be changed for each row, adding a column `digits` to the rifttable `design`:

```{r rounding_digits}
tribble(
  ~label,                     ~type,            ~digits,
  "Deaths/N",                 "outcomes/total", NA,  # Uses rifttable default
  "Risk",                     "risk",           NA,  # Uses risk_digits below
  "Risk ratio (95% CI)",      "",               NA,
  "  Rounded to 1 digit",     "rr",             1,
  "  Rounded to 2 digits",    "rr",             2,
  "Risk difference (95% CI)", "rd",             3) %>%  # Overrides risk_digits
  mutate(
    exposure = "stage",
    outcome = "death",
    effect_modifier = "receptor") %>%
rifttable(
  data = breastcancer,
  risk_digits = 1) %>%  # Fewer digits for risks, unless specified by "digits"
  rt_gt()  # obtain formatted output
```


# How can I model differences in continous outcomes?
# How can I create joint models?

The following two example use the `cancer` data from the survival package. To compare patient age by ECOG performance status, first recode the variable `ph.ecog` to be categorical (a `factor`), and exclude patients with rare elevated `ecog.ps`:
By default, regression models will be fit separately for each stratum of the `effect_modifier`. 

```{r setup}
data(cancer, package = "survival")
Append `"_joint"` to `"hr"`, `"rr"`, `"rd"`, `"irr"`, `"irrrob"`, `"diff"`, `"fold"`, `"foldlog"`, `"quantreg"`, or `"or"` to obtain "joint" models for exposure and effect modifier that have a single reference category.
 
Note that the joint model will be fit across all non-missing (`NA`) strata of the effect modifier, even if the `design` table does not request all strata be shown.

cancer <- cancer %>%
  filter(ph.ecog < 3) %>%
  mutate(ph.ecog = factor(ph.ecog))
Compare stratified models to joint models for risk differences (for simplicity of presentation, count data are omitted):

```{r joint}
tribble(
  ~label,                       ~type,      ~stratum,         
  "**Overall**",                "rd",       c("Low", "High"), 
  "",                           "",         "",               
  "**Stratified models**",      "",         "",               
  "  Low hormone receptor",     "rd",       "Low",           
  "  High hormone receptor",    "rd",       "High",           
  "",                           "",         "",               
  "**Joint models**",           "",         "",               
  "  Low hormone receptor",     "rd_joint", "Low",
  "  High hormone receptor",    "rd_joint", "High") %>%
  mutate(
    exposure = "stage",
    outcome = "death",
    effect_modifier = "receptor") %>%
  rifttable(data = breastcancer) %>%
  rt_gt()
```


Using `rifttable()`, one can calculate mean differences (`type = "diff"`) or quantile differences (`type = "quantreg"`, by default, at the median):

```{r table2_diff}
tribble(
  ~label,                ~type,
  "Observations",        "total",
  "Mean age",            "mean",
  "  Mean difference",   "diff",
  "Median age",          "median",
  "  Median difference", "quantreg") %>%
  mutate(
    exposure = "ph.ecog",
    outcome = "age") %>%
  rifttable(data = cancer) %>%
  rt_gt()  # obtain formatted output
```


# How can I model ratios of continuous outcomes?

Three types of ratios for continuous outcomes are implemented in `rifttable()`:

`type =` | Ratio of ... | Model | Use
-------|----------|-----------------------|-----------------------------
`"fold"` | Arithmetric means | Generalized linear model with Gaussian distribution and log link: `glm(y ~ x, family = gaussian(link = "log"))` | Tolerates `0` in the `outcome` variable. May be informative if outcome is normally distributed without transformation.
`"irrrob"` | Arithmetric means | Generalized linear model with Poisson distribution and log link: `glm(y ~ x, family = poisson())`, with robust (sandwich) standard errors | Tolerates `0` in the `outcome` variable. May be informative if outcome is normally distributed without transformation.
`"foldlog"` | Geometric means | Linear model with log-transformed outcome: `lm(log(y) ~ x)` | Does not tolerate `0` in the `outcome` variable, as `log(0)` is undefined (R returns `-Inf`). May be informative if outcome is normally distributed after log transformation.

In all models, after exponentiation, beta coefficients can be interpreted as ratios. rifttable automatically does all necessary transformations.

In the `cancer` data (see previous question), ratios of usual (arithmetic) means of age could be considered informative, given that `hist(cancer$age)` etc. does not show a major skew in this outcome.

```{r table2_ratio}
# How I do I change the level for confidence intervals?

Add a `ci` column to the `design`:

```{r change_ci}
tribble(
  ~label,                                       ~type,
  "Observations",                               "total",
  "Mean age",                                   "", 
  "  Arithmetic mean",                          "mean",
  "  Geometric mean",                           "geomean",
  "Mean ratio",                                 "",
  "  Ratio of arithmetric means",               "fold",
  "  Ratio of arithmetric means, empirical SE", "irrrob",
  "  Ratio of geometric means",                 "foldlog") %>%
  ~label,            ~type,                   ~ci,
  "Deaths/N (Risk)", "outcomes/total (risk)", NA,
  "Risk ratio",      "",                      NA,
  "  80% CI",        "rr",                    0.8,
  "  95% CI",        "rr",                    NA,  # Defaults to 0.95
  "  99% CI",        "rr",                    0.99) %>%
  mutate(
    exposure = "ph.ecog",
    outcome = "age") %>%
    exposure = "stage",
    outcome = "death") %>%
  rifttable(
    data = cancer, 
    ratio_digits = 3) %>%  # show extra digits to highlight (minor) differences
    data = breastcancer,
    risk_percent = TRUE) %>%
  rt_gt()  # obtain formatted output
```

Patients with ECOG performance status of 2, compared to those with performance status 0, are on average 8% older (or 9%, if based on geometric means).

# How do I make rifttable calculate an estimand that is not built-in?

While the package provides a number of estimators commonly used in epidemiology, it will never be able to include all possible estimators. However, any custom estimate can be integrated into a rifttable by a defining custom estimation function.

The subsequent example will reproduce the following basic rifttable, which shows the mean age by sex, stratified by ECOG performance status, in the `cancer` data set:

```{r custom, message = FALSE}
data(cancer, package = "survival")
cancer <- cancer %>%
  tibble::as_tibble() %>%
  mutate(
    sex = factor(
      sex,
      levels = 1:2,
      labels = c("Male", "Female")))

design <- tibble::tibble(
  type = "mean",
  exposure = "sex",
  outcome = "age",
  effect_modifier = "ph.ecog",
  stratum = 1:2,
  label = paste0("ECOG PS ", stratum, ": mean age"))

design %>%
  rifttable(
    data = cancer,
    overall = TRUE) %>%
  rt_gt()
```

Instead of relying on rifttable's built-in estimator `type = "mean"`, we will define a custom function that calculates the mean:

```{r custom_fn}
estimate_my_mean <- function(data, ...) {
  data %>%
    group_by(.exposure) %>%
    summarize(
      res = paste(
        round(
          mean(.outcome),
          digits = 3),
        "yrs"))
}
```

Use the custom function `my_mean` instead of the built-in `mean`:

```{r custom_use}
design %>%  # Edit the previous design
  mutate(
    type = "my_mean",  # Replace built-in "mean" by custom "my_mean"
    label = paste0(label, " (custom)")) %>%
  rifttable(
    data = cancer,
    overall = TRUE) %>%
  rt_gt()
```

Specifications for custom functions:

* The function name must start with `estimate_`; this prefix is to be omitted when later calling the custom function within `rifttable()`.
* The `data` provided to `rifttable()` will be available to the custom function as an argument named `data`.
* The `data` are already subsetted to the `stratum` of the `effect_modifier`, if applicable.
* Copies of key variables are accessible under the names, `.exposure`, `.outcome`, `.event`, `.time`, and `.time2`, as applicable.
* The function must accept all elements of a rifttable `design` (e.g., `confounders`,  `digits`, `na_rm`, etc.) and of the `rifttable()` function (e.g., `reference`, `risk_percent`) as arguments. Many may not be relevant and can be captured in the argument list, `...` (see example).
* The function must return a tibble/data frame with one column for `.exposure`, with one row per exposure category, and one string column `res` for the estimate.
