---
title: "Frequently Asked Questions"
output: 
  rmarkdown::html_vignette:
    df_print: kable
vignette: >
  %\VignetteIndexEntry{Frequently Asked Questions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Basic example

Most subsequent examples build on this simple rifttable:

```{r basic, message = FALSE}
library(rifttable)
library(dplyr)  # for data management, e.g., mutate()
data(breastcancer, package = "risks")

design <- tibble::tribble(
  ~label,                       ~type,                   ~stratum,         
  "Overall",                    "",                      "",               
  "– Deaths/N",                 "outcomes/total",        c("Low", "High"), 
  "– Risk",                     "risk",                  c("Low", "High"), 
  "– Risk ratio (95% CI)",      "rr",                    c("Low", "High"), 
  "– Risk difference (95% CI)", "rd",                    c("Low", "High"), 
  "",                           "",                      "",               
  "Low hormone receptor",       "",                      "",               
  "– Deaths/N (Risk)",          "outcomes/total (risk)", "Low",           
  "– Risk difference (95% CI)", "rd",                    "Low",           
  "High hormone receptor",      "",                      "",               
  "– Deaths/N (Risk)",          "outcomes/total (risk)", "High",
  "– Risk difference (95% CI)", "rd",                    "High") %>%
  mutate(
    exposure = "stage",
    outcome = "death",
    effect_modifier = "receptor")

rifttable(
  design = design,
  data = breastcancer)
```

# How do I add overall statistics?

Use the `overall` argument to show descriptive data for the entire `data` set.
Inferential estimators showing comparisons between exposure categories will be blank there.

```{r overall}
rifttable(
  design = design,
  data = breastcancer, 
  overall = TRUE)
```


# How do I test for trend?

Instead of *testing* a null hypothesis about a trend, **rifttable** proposes 
*estimating* the difference in the outcome for a one-unit higher exposure. 
This is also called a linear slope. Here, we estimate the risk associated for 
stage that is one category higher.

```{r trend}
rifttable(
  design = design %>%
    mutate(trend = "stage_numeric"),
  data = breastcancer %>%
    mutate(stage_numeric = as.numeric(stage)))
```


# How do I show multiple exposures in the same table?

Our simple toy dataset just has one exposure variable. For demonstration,
we just create a second variable, with two categories, "Level 1" and "Level 2,"
which is a simplified combination of the `stage` and `receptor` variables.

We will flip the table `layout` from `"rows"` (the default) to `"cols"` and 
concatenate two rifttables. We also need to give our new `exposure2` variable 
the same label as `stage` to make sure results appear in 
the same column.


```{r multiple}
breastcancer_2exposures <- breastcancer %>%
  mutate(
    exposure2 = case_when(
      stage == "Stage I" | 
        (stage == "Stage II" & receptor == "High") ~ 
        "Level 1",
      stage == "Stage III" | 
        (stage == "Stage II" & receptor == "Low") ~ 
        "Level 2"))

attr(breastcancer_2exposures$exposure2, which = "label") <- "Exposure"
attr(breastcancer_2exposures$stage, which = "label") <- "Exposure"

bind_rows(
  design %>%
    mutate(exposure = "exposure2") %>%
    slice(2:5) %>%
    rifttable(
      data = breastcancer_2exposures, 
      layout = "cols"),
  design %>%
    slice(2:5) %>%
    rifttable(
      data = breastcancer_2exposures, 
      layout = "cols")
)
```


# How do I change how results are rounded?

By default, differences measures are being rounded to 2 decimal digits (`0.01`), such as `type = "diff"`, the mean difference, or `type = "quantreg"`, the median difference. The same goes for risk measures, such as `type = "risk"`, unless shown as percentage points. Ratio measures are also shown with 2 decimal digits, such as `type = "hr"`, the hazard ratio, or `type = "fold"`, a ratio of arithmetric means. 

Rounding can be changed by setting the `rifttable()` arguments `diff_digits`, `risk_digits`, and `ratio_digits` globally for the entire table. 


```{r rounding}
design <- tibble::tribble(
  ~label,                     ~type,
  "Deaths/N",                 "outcomes/total", 
  "Risk",                     "risk",          
  "Risk ratio (95% CI)",      "rr", 
  "Odds ratio (95% CI)",      "or", 
  "Risk difference (95% CI)", "rd") %>%
  mutate(
    exposure = "stage",
    outcome = "death",
    effect_modifier = "receptor")

rifttable(
  design = design,
  data = breastcancer,
  ratio_digits = 3,  # Many digits for ratios
  risk_digits = 1)   # Fewer digits for risks
```

As can be seen, ratios > 3 are still shown with 1 fewer decimal, and ratios > 10 are shown with 2 fewer decimals ([Wilcox, *Epidemiology* 2004](https://doi.org/10.1097/01.ede.0000101026.08873.14) motivates why). To disable such additional rounding of extremely high ratios:

```{r rounding_decrease}
rifttable(
  design = design,
  data = breastcancer,
  ratio_digits = 3,
  ratio_digits_decrease = NULL,  # Do not round high ratios more
  risk_digits = 1)
```

Additionally, rounding can be changed for each row, adding a column `digits` to the rifttable `design`:

```{r rounding_digits}
tibble::tribble(
  ~label,                     ~type,            ~digits,
  "Deaths/N",                 "outcomes/total", NA,  # Uses rifttable default
  "Risk",                     "risk",           NA,  # Uses rifttable default
  "Risk ratio (95% CI)",      "",               NA,  # Uses risk_digits below
  "– Rounded to 1 digit",     "rr",             1,
  "– Rounded to 2 digits",    "rr",             2,
  "Risk difference (95% CI)", "rd",             3) %>%  # Overrides risk_digits
  mutate(
    exposure = "stage",
    outcome = "death",
    effect_modifier = "receptor") %>%
rifttable(
  data = breastcancer,
  risk_digits = 1)   # Fewer digits for risks, unless specified by "digits"

# How can I model differences in continous outcomes?

The following two example use the `cancer` data from the survival package. To compare patient age by ECOG performance status, first recode the variable `ph.ecog` to be categorical (a `factor`), and exclude patients with rare elevated `ecog.ps`:

```{r setup}
data(cancer, package = "survival")

cancer <- cancer %>%
  filter(ph.ecog < 3) %>%
  mutate(ph.ecog = factor(ph.ecog))
```


Using `rifttable()`, one can calculate mean differences (`type = "diff"`) or quantile differences (`type = "quantreg"`, by default, at the median):

```{r table2_diff}
tibble::tribble(
  ~label,                ~type,
  "Observations",        "total",
  "Mean age",            "mean",
  "  Mean difference",   "diff",
  "Median age",          "median",
  "  Median difference", "quantreg") %>%
  mutate(
    exposure = "ph.ecog",
    outcome = "age") %>%
  rifttable(data = cancer) %>%
  rt_gt()  # obtain formatted output
```


# How can I model ratios of continuous outcomes?

Three types of ratios for continuous outcomes are implemented in `rifttable()`:

`type =` | Ratio of ... | Model | Use
-------|----------|-----------------------|-----------------------------
`"fold"` | Arithmetric means | Generalized linear model with Gaussian distribution and log link: `glm(y ~ x, family = gaussian(link = "log"))` | Tolerates `0` in the `outcome` variable. Best fit if outcome is normally distributed without transformation.
`"irrrob"` | Arithmetric means | Generalized linear model with Poisson distribution and log link: `glm(y ~ x, family = poisson())`, with robust (sandwich) standard errors | Tolerates `0` in the `outcome` variable. (Note, `stats::glm()` may issue warnings that the outcome is "non-integer." This warning can be ignored here.)
`"foldlog"` | Geometric means | Linear model with log-transformed outcome: `lm(log(y) ~ x)` | Does not tolerate `0` in the `outcome` variable, as `log(0)` is undefined (R returns `-Inf`). Best fit if outcome is normally distributed after log transformation.

In all models, beta coefficients can be interpreted as ratios after exponentiation. `rifttable`()` automatically does all nessary transformations.

In the `cancer` data (see previous question), ratios of usual (arithmetric) means of age could be considered informative, given that `hist(cancer$age)` etc. does not show a major skew in this outcome.

```{r table2_ratio}
tibble::tribble(
  ~label,                                       ~type,
  "Observations",                               "total",
  "Mean age",                                   "mean",
  "  Ratio of arithmetric means",               "fold",
  "  Ratio of arithmetric means, empirical SE", "irrrob",
  "  Ratio of geometric means",                 "foldlog") %>%
  mutate(
    exposure = "ph.ecog",
    outcome = "age") %>%
  rifttable(
    data = cancer, 
    ratio_digits = 3) %>%  # show extra digits to highlight (minor) differences
  rt_gt()  # obtain formatted output
```

Patients with ECOG performance status of 2, compared to those with performance status 0, are on average 8% older, based on comparing arithmetric means (9%  based on geometric means).
