---
title: "A Descriptive Table 1"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A Descriptive Table 1}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Loading the package

After installation (using `remotes::install_github("stopsack/rifttable")`), load the package with:

```{r setup}
library(rifttable)
```


# Data handling

This example uses the `cancer` dataset from the survival package.

```{r data, message = FALSE}
library(dplyr) # for data handling
data(cancer, package = "survival")
cancer <- cancer %>%
  tibble::as_tibble() %>%
  dplyr::mutate(
    sex = factor(
      sex,
      levels = 1:2,
      labels = c("Male", "Female")
    ),
    ph.ecog = factor(ph.ecog)
  )
```


# Generate a table design

rifttable's `table1_design()` generates the `design` table that can then be passed on to `rifttable()`.

```{r design}
design <- cancer %>%
  table1_design(
    age, ph.ecog, ph.karno, pat.karno, # leave empty to include all variables
    by = sex
  )

# Print the design for illustration purposes
design
```


# Obtain Table 1

```{r rifttable}
design %>%
  rifttable() %>%
  rt_gt() # obtain formatted output
```

Of note, the dataset is passed along silently together with the `design`. However, the same design can be applied to different data, if it contains the same variable names, *e.g.*, `rifttable(design = design, data = otherdata)`.


# Customize the table

Label variables, set custom rounding, change estimands, and add "Overall" column.

```{r custom}
# Alternative: set_variable_labels() from the {labelled} package
attr(cancer$age, "label") <- "Age, years" # At diagnosis? Unclear (Loprinzi 94)
attr(cancer$ph.ecog, "label") <- "Physician-rated ECOG score"
attr(cancer$ph.karno, "label") <- "Physician-rated Karnofsky score"
attr(cancer$pat.karno, "label") <- "Patient-rated Karnofsky score"
attr(cancer$sex, "label") <- "Sex"

design <- cancer %>%
  table1_design(
    age, ph.ecog, ph.karno, pat.karno,
    by = sex,
    continuous_type = "mean (sd)"
  ) %>% # default: "median (iqr)"
  mutate( # rounding specifically for the "age" variable
    digits = if_else(
      outcome == "age",
      true = 1,
      false = NA
    )
  )

design %>%
  rifttable(
    diff_digits = 0, # rounding for continuous variables other than age
    overall = TRUE
  ) %>% # add unstratified "overall" column
  rt_gt() %>% # obtain formatted output
  gt::tab_footnote(
    footnote = "Data shown are count (percent) or mean (standard deviation)."
  )
```

# Add more statistics per variable

For the age variable, do not just display mean and standard deviation, but also the range. To this end, edit the `design` just like a regular dataset.

```{r morestats}
design_new <- design %>%
  mutate( # create three rows, not one, for "age"
    copy = if_else(
      outcome == "age",
      true = 3,
      false = 1
    )
  ) %>%
  tidyr::uncount(copy, .id = "copy") %>%
  mutate( # set new labels and types for the three rows on age
    label = case_when(
      outcome == "age" & copy == 2 ~ "  Mean (SD)",
      outcome == "age" & copy == 3 ~ "  Range",
      TRUE ~ label
    ),
    type = case_when(
      outcome == "age" & copy == 1 ~ "",
      outcome == "age" & copy == 2 ~ "mean (sd)",
      outcome == "age" & copy == 3 ~ "range",
      TRUE ~ type
    )
  )
design_new
design_new %>%
  rifttable(
    diff_digits = 0,
    overall = TRUE
  ) %>%
  rt_gt() %>%
  gt::tab_footnote(
    footnote = "Data shown are count (percent), unless indicated otherwise."
  )
```
